---
title: "Lab 2 — Part 4: Data Analytics with R and ClickHouse"
subtitle: "From Data Pipeline to Business Insights — v1.0.0"
author: "Allan Omondi"
date: today
date-format: "DD MMMM YYYY"

engine: knitr

format:
  html:
    toc: true
    toc-depth: 4
    toc-float: true
    number-sections: true
    fig-width: 6
    fig-height: 6
    embed-resources: true
    keep-md: false
    df-print: paged
    theme: flatly
    highlight-style: tango
  docx:
    toc: true
    toc-depth: 4
    number-sections: true
    fig-width: 6
    keep-md: false
  pdf:
    toc: true
    toc-depth: 4
    number-sections: true
    fig-width: 6
    fig-height: 6
    fig-crop: false
    keep-tex: false
    pdf-engine: xelatex

execute:
  echo: fenced
  message: false
  warning: false
  fig-width: 9
  fig-height: 5

knitr:
  opts_chunk:
    comment: "#>"
---

# Introduction

In Parts 1 through 3 we built a complete data engineering pipeline that moves
orders from a producer, through Kafka, into PostgreSQL, and then into ClickHouse
via Debezium CDC and a custom transformer service coded using Python.

In this notebook you will use R to:

1.  Connect to both databases and inspect the data
2.  Perform analytical queries to answer business questions
3.  Run the **same query** on PostgreSQL and ClickHouse and measure the
    performance difference
4.  Visualise the results using `ggplot2`

The central argument (the thesis) of this lab is simple: **the data pipeline we
built exists because analytical queries do not belong in the operational data
store (ACID-compliant relational databases designed to support database
transactions).** By the end of this notebook, you will:

1.  **Measure** the performance difference between executing an analytical query
    in an operational data store (relational \[object relational\] PostgreSQL
    database) versus in an analytical data store (ClickHouse non-relational
    \[column-oriented\] data warehouse).

--------------------------------------------------------------------------------

# Setup: Load Libraries and Connect

```{r setup_chunk}
#| echo: false
#| message: true
#| warning: true

knitr::opts_chunk$set(echo = TRUE)
# `installed.packages()` retrieves a matrix of all installed packages
# `[, "Package"]` extracts on the "Package" column from the matrix of all
# packages
# The %in% operator is used to test if the specified package is in the matrix of
# all packages
# `character.only = TRUE` ensures that the quoted name of the package is not treated
# as a symbol
# `dependencies = TRUE` instructs R to install not only the specified package but
# also its dependencies
# `pacman::p_load("here")` installs the package called "here". This package is used
# in the next line.
# `knitr::opts_knit$set(root.dir = here::here())` is used to ensure that the "knitr"
# utility in R knows where to find the files required to create the HTML, Word, or
# PDF version of the notebook.

if (!"pacman" %in% installed.packages()[, "Package"]) {
  install.packages("pacman", dependencies = TRUE)
  library("pacman", character.only = TRUE)
}

pacman::p_load("here")

knitr::opts_knit$set(root.dir = here::here())
```

We start by connecting to both PostgreSQL and ClickHouse using the `RPostgres`
package and the ClickHouse ODBC driver. The output confirms that the connection
is successful

```{r establish_connections}
#| echo: false
#| message: true
#| warning: true

# The code in this chunk is not displayed in the rendered output because of the
# plaintext password.

pacman::p_load("DBI")
pacman::p_load("RPostgres")
pacman::p_load("odbc")
pacman::p_load("tidyverse")
pacman::p_load("scales")
pacman::p_load("bench")

ch_conn <- DBI::dbConnect(
  odbc::odbc(),
  driver   = "ClickHouse ODBC Driver (Unicode)",
  Url      = "http://localhost:8123/",
  Username = "default",
  Password = "lab_password",
  Database = "default"
)

cat("✅ Connected to ClickHouse.\n")

pg_conn <- DBI::dbConnect(
  RPostgres::Postgres(),
  host     = "localhost",
  port     = 5432L,
  dbname   = "lab_db",
  user     = "lab_user",
  password = "lab_password"
)

cat("✅ Connected to PostgreSQL.\n")
```

```{r confirm_connections}
# These connections must already exist from the previous chunk.
stopifnot(
  "ch_conn not found — run connect_clickhouse_with_ODBC.R first" = exists("ch_conn"),
  "pg_conn not found — run connect_clickhouse_with_ODBC.R first" = exists("pg_conn")
)

# Confirm row counts
ch_count <- dbGetQuery(ch_conn, "SELECT COUNT(*) AS n FROM orders")$n
pg_count <- dbGetQuery(pg_conn, "SELECT COUNT(*) AS n FROM orders")$n

cat(sprintf("ClickHouse : %s rows\n", format(ch_count, big.mark = ",")))
cat(sprintf("PostgreSQL : %s rows\n", format(pg_count, big.mark = ",")))
```

--------------------------------------------------------------------------------

# Exploratory Data Analysis

The first step in any data-related project is to understand the structure and
distribution of the data before running statistical tests, training models,
solving optimization problems, creating dashboards, etc. This section
demonstrates that workflow using the ClickHouse data warehouse.

## Overview of the Orders Data

While `ROUND()` is excellent for final reports, it is often a mistake to apply
rounding within a query **if** those values are to be used for further
calculations later. Rounding early introduces "truncation error" that compounds
as you aggregate further. It is technically superior to keep the full precision
in the database and handle the visual formatting (the "two decimal places")
within the data analytics reporting layer itself using functions like `round()`
or `sprintf()`.

As noted in Part 3 of the lab, the timezone used in the PostgreSQL database is
UTC+2 (Lagos, Nigeria) then the transformer converts this so that by the time it
is stored in the ClickHouse data warehouse, it is stored as UTC+3 (Nairobi,
Kenya).

In other words, consider a multinational business. PostgreSQL is in a branch in
Lagos, Nigeria while ClickHouse is in the Headquarters in Nairobi, Kenya.

So the data retrieved from ClickHouse is already in UTC+3 or East African Time
(EAT) thanks to the work done by the transformer.

```{r eda-overview}
# Pull a representative sample for local exploration.
# Note: for summarising the full dataset, we push computation to ClickHouse
# using SQL and only bring the result into R. This is best practice —
# never pull millions of rows into R when a database can compute the 
# summary for you. Otherwise, it beats the purpose of the business investing in
# acquiring a powerful database server or data warehouse or data lake.
overview <- dbGetQuery(ch_conn, "
  SELECT
    COUNT(*)                               AS total_orders,
    COUNT(DISTINCT customer_name)          AS unique_customers,
    COUNT(DISTINCT item)                   AS unique_items,
    SUM(order_quantity)                    AS total_units,
    ROUND(AVG(order_quantity), 2)          AS avg_quantity_per_order,
    SUM(is_bulk_order)                     AS bulk_orders,
    ROUND(SUM(is_bulk_order) / COUNT(*) * 100, 2) AS bulk_order_pct,
    toTimeZone(MIN(received_at), 'Africa/Nairobi') AS earliest_order,
    toTimeZone(MAX(received_at), 'Africa/Nairobi') AS latest_order
  FROM orders
  WHERE operation != 'DELETE'
")

overview$total_orders            <- as.integer(overview$total_orders)
overview$unique_customers        <- as.integer(overview$unique_customers)
overview$unique_items            <- as.integer(overview$unique_items)
overview$total_units             <- as.integer(overview$total_units)
overview$avg_quantity_per_order  <- as.numeric(overview$avg_quantity_per_order)
overview$bulk_orders             <- as.integer(overview$bulk_orders)
overview$bulk_order_pct          <- as.numeric(overview$bulk_order_pct)

overview$earliest_order <- as.POSIXct(overview$earliest_order, tz = "UTC")
overview$latest_order   <- as.POSIXct(overview$latest_order, tz = "UTC")

overview$earliest_order <- format(
  overview$earliest_order,
  "%Y-%m-%d %H:%M:%S %Z",
  tz = "Africa/Nairobi"
)

overview$latest_order <- format(
  overview$latest_order,
  "%Y-%m-%d %H:%M:%S %Z",
  tz = "Africa/Nairobi"
)

overview_t <- data.frame(
  Metric = names(overview),
  Value  = I(as.list(overview[1, ]))
)

overview_t$Value <- sapply(overview_t$Value, function(x) {
  if (inherits(x, "POSIXct")) {
    format(x, "%Y-%m-%d %H:%M:%S %Z")
  } else if (is.numeric(x)) {
    format(x, scientific = FALSE, big.mark = ",")
  } else {
    as.character(x)
  }
})

print(overview_t)
```

**Reading the table:** The `bulk_order_pct` shows what proportion of all orders
are classified as bulk (quantity \> 5). The gap between `earliest_order` and
`latest_order` shows the time window covered by the dataset. These are possible
first questions a data analyst can ask when handed a new dataset.

## Order Distribution by Item

```{r orders-by-item}
item_summary <- dbGetQuery(ch_conn, "
  SELECT
    item,
    COUNT(*)          AS order_count,
    SUM(order_quantity) AS total_units,
    AVG(order_quantity) AS avg_quantity,
    SUM(is_bulk_order)  AS bulk_orders
  FROM orders
  WHERE operation != 'DELETE'
  GROUP BY item
  ORDER BY total_units DESC
")

pacman::p_load("ggplot2")
pacman::p_load("scales")

# `x = reorder(item, total_units)` Reorders items by total_units
# The default is to order in alphabetical order which is rarely useful for analysis
ggplot(item_summary, aes(x = reorder(item, total_units), y = total_units)) +
  geom_col(fill = "#4F6EA4", width = 0.7) +
  geom_text(
    # Explicitly cast to numeric to ensure format() recognizes the number
    aes(label = format(as.numeric(total_units), big.mark = ",")), 
    hjust = -0.2, 
    size = 3.2
  ) +
  # `coord_flip()` converts the standard vertical bar chart to a horizontal bar chart
  coord_flip() +
  scale_y_continuous(
    labels = label_comma(), 
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title    = "Total Units Ordered by Item",
    subtitle = "Aggregated across all orders in the ClickHouse data warehouse",
    x        = "Item",
    y        = "Total Units",
    caption  = "Source: ClickHouse — orders table"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold"),
    panel.grid.major.y = element_blank()
  )
```

## Bulk versus Standard Order Breakdown

```{r bulk-vs-standard}
bulk_summary <- dbGetQuery(ch_conn, "
  SELECT
    item,
    SUM(CASE WHEN is_bulk_order = 1 THEN 1 ELSE 0 END) AS bulk,
    SUM(CASE WHEN is_bulk_order = 0 THEN 1 ELSE 0 END) AS standard
  FROM orders
  WHERE operation != 'DELETE'
  GROUP BY item
  ORDER BY item
")

# Reshape to long format for grouped bar chart
bulk_long <- bulk_summary %>%
  pivot_longer(cols = c(bulk, standard),
               names_to  = "order_type",
               values_to = "count") %>%
  mutate(order_type = str_to_title(order_type))

ggplot(bulk_long, aes(x = reorder(item, -count), y = count, fill = order_type)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(
    values = c("Bulk" = "#E03C31", "Standard" = "#4F6EA4"),
    name   = "Order Type"
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Bulk vs Standard Orders by Item",
    subtitle = "Bulk → order quantity > 5 units",
    x        = "Item",
    y        = "Number of Orders",
    caption  = "Source: ClickHouse — orders table"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold"),
    legend.position    = "bottom",
    axis.text.x        = element_text(angle = 30, hjust = 1)
  )
```

## Order Volume Over Time

```{r orders-over-time}
# Aggregate orders by week to show trends over the 90-day period
weekly_orders <- dbGetQuery(ch_conn, "
  SELECT
    toStartOfWeek(received_at)   AS week_start,
    COUNT(*)                     AS order_count,
    SUM(order_quantity)          AS total_units
  FROM orders
  WHERE operation != 'DELETE'
  GROUP BY week_start
  ORDER BY week_start
")

weekly_orders$week_start <- as.Date(weekly_orders$week_start)

ggplot(weekly_orders, aes(x = week_start, y = order_count)) +
  geom_line(colour = "#4F6EA4", linewidth = 1.2) +
  geom_point(colour = "#E03C31", size = 2.5) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 weeks") +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Weekly Order Volume",
    subtitle = "90-day view from ClickHouse warehouse data",
    x        = "Week Starting",
    y        = "Number of Orders",
    caption  = "Source: ClickHouse — orders table"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold"),
    axis.text.x  = element_text(angle = 30, hjust = 1)
  )
```

--------------------------------------------------------------------------------

# Hypothesis Tests

A data analyst does not simply describe data and create appealing
visualizations. [**No!**]{.underline} They proceed to create hypotheses and test
these hypotheses. The answers to the hypothesis tests is what supports
**data-driven decisions**. This section demonstrates two statistical tests using
the orders data. Subsequent labs will focus on other statistical tests of
hypothesis that you need to learn how to perform.

## Hypothesis Test 1: Do Bulk Orders Differ in Quantity?

**Business question:** Do customers who place bulk orders (quantity \> 5) place
significantly larger orders than customers who place standard orders? Or is the
5-unit threshold arbitrary?

**Hypotheses:**

- H₀: The mean order quantity for bulk orders equals the mean order quantity for
  standard orders. (μ_bulk = μ_standard)
- H₁: The mean order quantity for bulk orders is greater than for standard
  orders. (μ_bulk \> μ_standard)

```{r ttest-bulk}
# Pull the raw quantities — we need individual records for the t-test
quantities <- dbGetQuery(ch_conn, "
  SELECT order_quantity, is_bulk_order
  FROM orders
  WHERE operation != 'DELETE'
  LIMIT 200000
")

bulk_qty     <- quantities$order_quantity[quantities$is_bulk_order == 1]
standard_qty <- quantities$order_quantity[quantities$is_bulk_order == 0]

cat(sprintf("Bulk orders     : n = %s, mean = %.2f, sd = %.2f\n",
    format(length(bulk_qty), big.mark=","),
    mean(bulk_qty), sd(bulk_qty)))

cat(sprintf("Standard orders : n = %s, mean = %.2f, sd = %.2f\n",
    format(length(standard_qty), big.mark=","),
    mean(standard_qty), sd(standard_qty)))

# One-sided Welch's t-test (does not assume equal variances)
t_result <- t.test(
  bulk_qty,
  standard_qty,
  alternative = "greater",
  var.equal   = FALSE
)

print(t_result)

cat("\n--- INTERPRETATION ---\n")
if (t_result$p.value < 0.05) {
  cat(sprintf(
    # Note: We use %e (not e.g., %.6f) because the p-value is significantly
    # smaller than 0.0001
    "We reject H₀ at the 5%% significance level (p = %e).\n",
    t_result$p.value
  ))
  cat("There is statistically significant evidence that bulk orders\n")
  cat("have a higher mean quantity than standard orders.\n")
} else {
  cat(sprintf(
    "We fail to reject H₀ (p = %e).\n",
    t_result$p.value
  ))
  cat("There is insufficient evidence of a significant difference.\n")
}
```

```{r ttest-visualise}
# Visualise the distribution of quantities for both groups
quantities %>%
  mutate(order_type = if_else(is_bulk_order == 1, "Bulk", "Standard"),
         order_type = factor(order_type, levels = c("Standard", "Bulk"))) %>%
  ggplot(aes(x = order_quantity, fill = order_type)) +
  # When you use `geom_histogram(binwidth = 1)`, R centers the bins over the
  # integers by default. This means:
    # The bin for a quantity of 5 actually covers the range from 4.5 to 5.5.
    # The bin for a quantity of 6 actually covers the range from 5.5 to 6.5.
  # If we were to place your xintercept exactly at 5.0, the line would cut right
  # through the middle of the "Standard" bin for 5 units.
  # By placing it at 5.5, we are drawing the line in the "no man's land" between
  # the bin for 5 and the bin for 6. This visually reinforces the logical boundary
  # that anything greater than 5 (starting at 6) is considered "Bulk."
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.65) +
  geom_vline(
    xintercept = 5.5, linetype = "dashed",
    colour = "#1D1D1D", linewidth = 0.8
  ) +
  annotate("text", x = 6.2, y = Inf, vjust = 5.5,
           label = "Bulk threshold (>5)", size = 3.5) +
  scale_fill_manual(
    values = c("Standard" = "#4F6EA4", "Bulk" = "#E03C31"),
    name   = "Order Type"
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Distribution of Order Quantities",
    subtitle = "Bulk orders (quantity > 5) vs Standard orders",
    x        = "Order Quantity",
    y        = "Count",
    caption  = "Source: ClickHouse — 200,000 row sample"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold"),
    legend.position = "top"
  )
```

## Hypothesis Test 2: Is Order Quantity Independent of Item Type?

**Business question:** Does the type of item ordered influence whether it is a
bulk or standard order? Or do customers order all items in roughly equal
proportions of bulk and standard?

**Hypotheses:**

- H₀: Order type (bulk vs standard) is independent of item type.
- H₁: There is a significant association between item type and order type.

```{r chisq-test}
# Build a contingency table
contingency_data <- dbGetQuery(ch_conn, "
  SELECT
    item,
    SUM(CASE WHEN is_bulk_order = 1 THEN 1 ELSE 0 END) AS bulk,
    SUM(CASE WHEN is_bulk_order = 0 THEN 1 ELSE 0 END) AS standard
  FROM orders
  WHERE operation != 'DELETE'
  GROUP BY item
  ORDER BY item
")

# Explicitly cast to integer to clear the bit64 mapping issue common in ODBC
contingency_data$bulk <- as.integer(contingency_data$bulk)
contingency_data$standard <- as.integer(contingency_data$standard)

# Disable scientific notation (e.g., 7.156541e-320) for the current session
options(scipen = 999)

# Convert to matrix for chi-square test
contingency_matrix <- as.matrix(contingency_data[, c("bulk", "standard")])
rownames(contingency_matrix) <- contingency_data$item

cat("Contingency table:\n")
print(contingency_matrix)

# Run the test
chi_result <- chisq.test(contingency_matrix)
print(chi_result)

cat("\n--- INTERPRETATION ---\n")

# -----------------------
# Assumption Check: Ensure all expected frequencies are >= 5
expected <- chi_result$expected

if (any(expected < 5)) {
  warning("Some expected counts are below 5. Chi-square assumptions violated.")
}

# -----------------------
# Final Hypothesis Decision
if (is.na(chi_result$p.value)) {
  cat("Chi-square test could not compute a valid p-value.\n")
  cat("This usually happens when expected frequencies are too small\n")
  cat("or when some categories have zero counts.\n")
  cat("Consider combining categories or using Fisher's Exact Test in this case.\n")
} else if (chi_result$p.value < 0.05) {
  cat(sprintf(
    "We reject H₀ at the 5%% significance level (p = %e).\n",
    chi_result$p.value
  ))
  cat("There is a statistically significant association between\n")
  cat("item type and whether an order is bulk or standard.\n")
  cat("In summary: The item choice influences the likelihood of a bulk order.\n")
  cat("\nBoardroom-Ready Summary (plain language): The type of food ordered\n",
      "(e.g., Ugali vs. Managu) directly influences whether an order is\n",
      "likely to be Bulk or Standard. They are not independent; certain items\n",
      "are much more prone to being ordered in large quantities than others.")
} else {
  cat(sprintf(
    "We fail to reject H₀ (p = %.4f).\n",
    chi_result$p.value
  ))
  cat("Item type and order type appear to be independent.\n")
  cat("In summary: The item choice does not seem to affect order size.\n")
}
```

--------------------------------------------------------------------------------

# The Performance Benchmark

This is the central justification for the data pipeline you built. You will now
run the **same aggregation query** on PostgreSQL and on ClickHouse and measure
the execution time of each.

## The "Cold Start" Problem

The "Cold Start" problem in this context of database benchmarking describes
**the performance lag that occurs when a system is first initialized or
restarted, forcing it to fetch data from the physical disk storage rather than
from high-speed memory (RAM)**.

### The Concept: Cold versus Warm

- **Cold Start:** The database has just started. Its memory (cache) is empty.
  Every request for data requires a "trip" to the hard drive or SSD, which is
  significantly slower than RAM.

- **Warm Start:** After the first query, the database stores the results and
  frequently accessed data in its RAM. Subsequent queries are "warm" because the
  system can skip the disk trip and serve data instantly from memory.

### Library Analogy

Imagine you are a researcher working on your final year research and you need
literature from a book available in the library:

- **Cold Start:** You enter the library for the first time today. To find
  information on "Test Driven Development (TDD)", you must walk deep into the
  aisles on a specific floor, find the specific shelf, and pull out the book.
  This takes ≈10 minutes.

- **Warm Start:** You already found the book on TDD. You leave the book open on
  your desk as you continue to research on a similar topic about Requirements
  Driven Development (RDD). The next time you need to check a detail on TDD
  again, you simply look down at the table where you left the book. This takes
  ≈2 seconds.

### Why it Matters for Benchmarking

If you do not restart ("Cold Start") the server (the Docker containers),
PostgreSQL might appear as fast as ClickHouse simply because it is reading from
the "desk" (RAM). By forcing a Cold Start, we are testing the **architectural
efficiency** of the database engines—specifically how ClickHouse’s columnar
storage allows it to scan only the necessary data from the disk, whereas
PostgreSQL’s row-based structure forces it to read much more than is actually
required.

### Strategic Counter-Position

While Cold Start benchmarks reveal raw architectural power, **Warm Start**
benchmarks are arguably more representative of a healthy production environment.
An optimized PostgreSQL instance should ideally be "warm" 99% of the time. The
real "victory" for ClickHouse is not just that it is faster on a cold start, but
that its columnar compression allows it to keep **much more data** in a "warm"
state using the same amount of RAM compared to PostgreSQL.

Run this in your terminal:

`cd 3_data_engineering/docker restart postgres clickhouse && docker ps`

Wait until both PostgreSQL and ClickHouse are ready to receive connections. You
can confirm that they are ready by connecting via DataGrip.

## The Query

The benchmark query computes, for each item, the total number of orders, the
total units ordered, the average order quantity, and the bulk order rate. This
is a typical OLAP query — it scans every row, reads multiple columns, and
produces a grouped aggregation.

```{r benchmark-query-text}
# Define the queries for each database.
# The logic is identical — only the column names differ (client_fname vs
# customer_name) because the transformer renamed the field.

PG_QUERY <- "
  SELECT
    item,
    COUNT(*)                              AS order_count,
    SUM(order_quantity)                   AS total_units,
    ROUND(AVG(order_quantity), 2)         AS avg_quantity,
    SUM(CASE WHEN order_quantity > 5 
             THEN 1 ELSE 0 END)           AS bulk_orders,
    ROUND(((SUM(CASE WHEN order_quantity > 5 THEN 1 ELSE 0 END))::numeric / COUNT(*)) * 100, 2) AS bulk_pct
  FROM orders
  GROUP BY orders.item
  ORDER BY total_units DESC
"

CH_QUERY <- "
  SELECT
      item,
      COUNT(*)                              AS order_count,
      SUM(order_quantity)                   AS total_units,
      ROUND(AVG(order_quantity), 2)         AS avg_quantity,
      SUM(CASE WHEN order_quantity > 5 
               THEN 1 ELSE 0 END)           AS bulk_orders,
      ROUND(CAST((SUM(CASE WHEN order_quantity > 5 THEN 1 ELSE 0 END)) AS Float64) / COUNT(*) * 100, 2) AS bulk_pct
    FROM orders
    WHERE orders.operation != 'DELETE'
    GROUP BY orders.item
    ORDER BY total_units DESC
"

cat("Both queries are defined and ready.\n")
cat("PostgreSQL query will scan:  ", format(pg_count, big.mark=","), " rows\n")
cat("ClickHouse query will scan:  ", format(ch_count, big.mark=","), " rows\n")
```

## Run the Benchmark

The `bench::mark()` function runs each query multiple times and reports the
median, mean, and standard deviation of execution times. Using `bench` rather
than `system.time()` produces more reliable results because it controls for CPU
caching and Just-In-Time (JIT) effects.

Normally, code is either **interpreted** (read line-by-line, which is slow) or
**compiled** (translated into machine code all at once, which is fast). JIT is a
hybrid. The system starts by interpreting the code, but as it notices certain
parts are being run frequently, it compiles those specific parts into optimized
machine code "on the fly" (Just-In-Time).

```{r benchmark-run}
# Warm up both connections with a trivial query before benchmarking.
# This ensures the connection overhead is not included in the timed results.
invisible(dbGetQuery(pg_conn, "SELECT 1"))
invisible(dbGetQuery(ch_conn, "SELECT 1"))

cat("Running benchmark — this may take 1 to 2 minutes...\n")
cat("Each query runs", 10, "times. Median time is reported.\n\n")

benchmark_results <- bench::mark(
  PostgreSQL  = dbGetQuery(pg_conn, PG_QUERY),
  ClickHouse  = dbGetQuery(ch_conn, CH_QUERY),
  iterations  = 10,
  check       = FALSE
)

# Extract median times for comparison
pg_median <- as.numeric(benchmark_results$median[1])
ch_median <- as.numeric(benchmark_results$median[2])
speedup   <- pg_median / ch_median

cat("\n===== BENCHMARK RESULTS =====\n")
cat(sprintf(
  "PostgreSQL  median : %s\n",
  format(benchmark_results$median[1])
))
cat(sprintf(
  "ClickHouse  median : %s\n",
  format(benchmark_results$median[2])
))
cat(sprintf(
  "ClickHouse was %.1f× faster than PostgreSQL\n",
  speedup
))
cat("=============================\n")
```

## Visualise the Results

```{r benchmark-visualise}
# Build a data frame from the benchmark results for plotting
bench_df <- data.frame(
  database       = c("PostgreSQL", "ClickHouse"),
  median_seconds = c(pg_median, ch_median),
  db_type        = c("OLTP (Row Store)", "OLAP (Column Store)")
)

ggplot(bench_df,
       aes(x = reorder(database, -median_seconds),
           y = median_seconds,
           fill = database)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(
    aes(label = sprintf("%.3f s", median_seconds)),
    vjust = -0.5, size = 4.5, fontface = "bold"
  ) +
  geom_text(
    aes(label = db_type),
    vjust = 1.8, size = 3.5, colour = "white", fontface = "italic"
  ) +
  scale_fill_manual(values = c(
    "PostgreSQL"  = "#E03C31",
    "ClickHouse"  = "#4F6EA4"
  )) +
  scale_y_continuous(
    labels   = label_number(suffix = " s"),
    expand   = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title    = "Aggregation Query: PostgreSQL vs ClickHouse",
    subtitle = sprintf(
      "Similar GROUP BY query run on %s rows. ClickHouse is ≈%.1f× faster",
      format(as.numeric(max(pg_count, ch_count)), big.mark = ","),
      speedup
    ),
    x        = NULL,
    y        = "Median Execution Time (seconds)",
    caption  = paste(
      "Benchmark: 10 iterations, using `bench::mark()`",
      sprintf("| %s rows",
              format(as.numeric(max(pg_count, ch_count)), big.mark = ","))
    )
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "#1D1D1D"),
    axis.text.x   = element_text(size = 12, face = "bold")
  )
```

## Examine the Query Results

Both queries return a similar business answer — only via different paths.

```{r benchmark-results-table}
pg_result <- dbGetQuery(pg_conn, PG_QUERY)
ch_result <- dbGetQuery(ch_conn, CH_QUERY)

knitr::kable(
  pg_result,
  caption   = "Aggregation Result from PostgreSQL",
  col.names = c("Item", "Order Count", "Total Units",
                "Avg Quantity", "Bulk Orders", "Bulk %"),
  format.args = list(big.mark = ",")
)
```

```{r}
knitr::kable(
  ch_result,
  caption   = "Aggregation Result from ClickHouse",
  col.names = c("Item", "Order Count", "Total Units",
                "Avg Quantity", "Bulk Orders", "Bulk %"),
  format.args = list(big.mark = ",")
)
```

## Why the Gap Exists: Column Store vs Row Store

```{r architecture-explanation}
cat(
"HOW POSTGRESQL EXECUTES THIS QUERY\n",
"─────────────────────────────────────────────────────────────────────────\n",
"1. Reads every row from disk (all columns: order_id, client_fname,\n",
"   item, order_quantity, received_at).\n",
"2. For each row, extracts only item and order_quantity.\n",
"3. Groups the extracted values and computes the aggregation.\n",
"4. The full row width is read even though only 2 of 5 columns are needed.\n",
"\n",
"HOW CLICKHOUSE EXECUTES THIS QUERY\n",
"─────────────────────────────────────────────────────────────────────────\n",
"1. Reads ONLY the columns referenced in the query:\n",
"   item, order_quantity, is_bulk_order.\n",
"2. These columns are stored contiguously on disk — no row crossing.\n",
"3. ClickHouse also compresses each column independently (similar values\n",
"   compress extremely well — e.g., repeated item names).\n",
"4. The grouped aggregation runs on the compressed column data directly.\n",
"\n",
"The ratio you measured is not an optimisation trick or a lucky cache hit.\n",
"It is a structural consequence of how the data is physically laid out on\n",
"disk. At 10 million rows this gap widens. At 1 billion rows it becomes\n",
"the difference between a 3-second dashboard refresh and a 45-minute wait.\n",
sep = ""
)
```

--------------------------------------------------------------------------------

# Data Pipeline Latency Analysis

One of the audit fields the transformer added (feature engineering) is
`processed_at` — the Nairobi time at which the transformer consumed the Kafka
message and wrote the record to ClickHouse. The gap between `received_at` and
`processed_at` is the end-to-end pipeline latency.

```{r latency}
latency_data <- dbGetQuery(ch_conn, "
  SELECT
    item,
    operation,
    AVG(
      dateDiff('second', received_at, processed_at)
    ) AS avg_latency_seconds,
    MAX(
      dateDiff('second', received_at, processed_at)
    ) AS max_latency_seconds,
    COUNT(*) AS sample_size
  FROM orders
  WHERE operation IN ('INSERT', 'SNAPSHOT')
    AND processed_at > received_at
  GROUP BY item, operation
  ORDER BY avg_latency_seconds DESC
")

knitr::kable(
  latency_data,
  caption   = "Average Pipeline Latency by Item and Operation Type",
  col.names = c("Item", "Operation", "Avg Latency (s)",
                "Max Latency (s)", "Sample Size"),
  format.args = list(big.mark = ","),
  digits = 2
)
```

```{r latency-visualise}
latency_data %>%
  filter(sample_size > 100) %>%
  ggplot(aes(x = reorder(item, avg_latency_seconds),
             y = avg_latency_seconds,
             fill = operation)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(
    values = c("INSERT" = "#4F6EA4", "SNAPSHOT" = "#CEA152"),
    name   = "Operation"
  ) +
  labs(
    title    = "Average End-to-End Pipeline Latency",
    subtitle = "Time between PostgreSQL write (received_at) and\nClickHouse write (processed_at)",
    x        = NULL,
    y        = "Average Latency (seconds)",
    caption  = "Source: ClickHouse — orders table | Only items with >100 records shown"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold"),
    panel.grid.major.y = element_blank(),
    legend.position    = "top"
  )
```

**Interpretation:** Snapshot records (op = 'r') typically show higher latency
than live inserts because the snapshot runs once at connector stars and
processes a large batch of existing records. Live inserts flow through the
pipeline in near real-time. A latency of under 5 seconds from PostgreSQL write
to ClickHouse availability is consistent with a well-functioning CDC pipeline.

--------------------------------------------------------------------------------

# Conclusion

This notebook was used to demonstrated three things:

**1. A data warehouse designed using a non-relational data model (ClickHouse)
enables analytical queries that a database designed using a relational (object
relational) data model (PostgreSQL) cannot serve at scale.** The benchmark
showed a clear performance difference on the same query against the same data
volume. This difference grows with scale.

**2. Statistical analysis answers real business questions.** The t-test
confirmed that bulk orders genuinely differ in quantity from standard orders.
The chi-square test revealed whether item type influences ordering behaviour.
These are questions a data analyst would ask.

**3. The pipeline latency is measurable and meaningful.** The `processed_at`
audit field, added by the transformer, enables pipeline observability — you can
measure how long it takes for a PostgreSQL change to appear in the warehouse.

The full architecture you have built across Part 1 through 4 — from Kafka
producer to ClickHouse analytics — is a simplified but architecturally faithful
representation of how data platforms are built in production environments.

--------------------------------------------------------------------------------

# Clean Up

Always close database connections when you are finished.

```{r cleanup}
dbDisconnect(ch_conn)
dbDisconnect(pg_conn)

cat("✅ Both connections closed.\n")
```
